---
title: Two-minute plot - county-level drought impact categories
author: ~
date: '2020-04-05'
slug: more-drought-data
categories: []
tags: []
description: County-level drought impact map in two minutes
---
The University of Nebraska Lincoln maintains an excellent [United States Drought Monitor](https://droughtmonitor.unl.edu) providing drought impact categories at various levels of granularity.

Here's a demo for quickly loading, wrangling, and plotting this week's county-level data.

First we need some libraries...

```{r, message = FALSE}
library(vroom)   # for data reading
library(dplyr)   # for data wrangling
library(ggplot2) # for plotting
library(viridis) # for nice color scheme
```

Then we need to read the data. The county-level data gives % land area within each drought category. Here I'm just going to use `dplyr::case_when` to select the most severe impact category present in each county.
```{r, message = FALSE}
vroom("https://droughtmonitor.unl.edu/Data/GISData.aspx?mode=table&aoi=county&date=") %>% 
  mutate(worst_case_impact = case_when(
    D4 > 0 ~ "D4",
    D3 > 0 ~ "D3",
    D2 > 0 ~ "D2",
    D1 > 0 ~ "D1",
    TRUE ~ "Neutral"
  )) %>% 
  select(fips = FIPS, worst_case_impact) %>% 
  mutate(fips = as.integer(fips)) ->
  county_impact
```

Then I'm gonna read some county polygon data using the `map_data` function of `ggplot2`:
```{r, message = FALSE}
map_data("county",
         projection = "albers",
         parameters = c(39, 45)) %>% 
  as_tibble() %>% 
  rename(state = region, county = subregion) ->
  county_data
```

My county name formats are not consistent--so I need some additional info to connect the drought data to the polygons. I'm gonna use a dataset in the `maps` package to get county FIPS code that will allow for a clean join:

```{r, message = FALSE}
maps::county.fips %>% 
  as_tibble() %>% 
  tidyr::separate(polyname, into = c("state", "county"), sep = ",") ->
  fips_state_county

county_data %>% 
  left_join(fips_state_county) %>% 
  left_join(county_impact) ->
  county_plot
```

And now for the plot.

```{r, message = FALSE}

# US state boundaries
map_data("state",
         projection = "albers",
         parameters = c(39, 45)) ->
  state_boundaries

# color scheme
c(viridis(4, option = "A", direction = -1,
          begin = 0.4, end = 0.9),
  "darkgrey") -> drought_cols

# main plot
county_plot %>% 
  ggplot(aes(long, lat, group = group)) +
  geom_polygon(aes(fill = worst_case_impact),
               colour = alpha("white", 1 / 2),
               size = 0.2) +
  geom_polygon(data = state_boundaries,
               colour = "white", fill = NA) +
  coord_fixed() +
  theme_void() +
  labs(title = "US drought impact status April 2020",
       fill = "Impact category") +
  scale_fill_manual(values = drought_cols)
```


