---
title: Glacier data wrangle
author: Sean Turner
date: '2020-05-04'
slug: a-look-at-glacier-projections
categories: []
tags: []
description: Plotting the new Bosson et al. data
---

Shout out to Jean Baptiste Bosson and colleagues for their recent paper in Earth's Future, complete with a nice open dataset of global glacier area and volume projections.

Let's wrangle 'em in `R`.

Note that I've downloaded two datasets for this analysis:

###### 1. [US National Park boundary shapefile](https://public-nps.opendata.arcgis.com/datasets/nps-boundary-1)

###### 2. [Bosson et al. Natural World Heritage Site Glacier projections](https://agupubs.onlinelibrary.wiley.com/doi/full/10.1029/2018EF001139)


First step: load some libraries...
```{r, message = F}
library(sf)      # for shape files
library(vroom)   # for .dat files
library(tmap)    # for map plotting
library(dplyr)   # for wrangling
library(purrr)   # for mapping through files
library(ggplot2) # for plotting
```

Next I'm loading two shape files...

```{r, message = F}
# Glacier for Olympic National Park
st_read(paste0(
  "../../../Large data files/",
  "Earths_Future_2019_bosson_al_data/",
  "RGI6.0 shapefiles of Glaciers in Natural World Heritage Sites in 2018/",
  "19_Olympic_National_Park.shp"
  ), quiet = TRUE) -> glacier_shape

# Shapefile for National Park boundaries
st_read(paste0(
  "../../../Large data files/",
  "NPS_-_Land_Resources_Division_Boundary_and_Tract_Data_Service/",
  "622305c9-fb01-40e6-97ac-f68aca551c38202034-1-xpq6gt.ugbbb.shp"
  ), quiet = TRUE) %>%
  subset(PARKNAME == "Olympic") ->
  Olympic_np

# test for same coordinate reference system?
st_crs(glacier_shape) == st_crs(Olympic_np)


# plot
tm_shape(Olympic_np) +
  tm_polygons(col = "lightgrey") +
  tm_shape(glacier_shape) +
  tm_polygons(col = "blue",
              border.col = NULL) +
  tm_layout(frame = FALSE,
            title = "Olympic National Park Glaciers")

```

So we can see some limited glaciation in Olympic National Park.

What's going to happen to those glaciers this century?

Now for the Bosson et al. simulations...

```{r}
# get file location
paste0(
  "../../../Large data files/",
  "Earths_Future_2019_bosson_al_data/",
  "Glacier Modelling by World Heritage Site/",
  "Olympic_National_Park/19_total_Volume_"
  ) -> file_loc
  
# map through available rcps
c("rcp26", "rcp45", "rcp85") %>% 
  map_dfr(function(rcp){
    # this part is a little funky...
    # ... cos the header and data delims are...
    # ... inconsistent in these .dat files !
    vroom(paste0(file_loc, rcp, ".dat"),
          skip = 1, col_names = F,
          delim = "       ", col_types = cols()) %>% 
      select(year = X1, mean = X2, sd = X3) %>% 
      mutate(rcp = !!rcp)
  }) -> glac_volume_series

# wrangle for ribbon plot
glac_volume_series %>% 
  mutate(max_ = mean + sd, min_ = mean - sd) %>% 
  ggplot(aes(x = year, fill = rcp)) +
  geom_ribbon(aes(ymax = max_, ymin = min_),
              alpha = 0.5) +
  theme_classic() +
  labs(title = "Volume of glaciers in Olympic National Park (cubic km)",
       x = NULL)
```

Bye bye Olympic National Park glaciers :(

May the 4th be with you.


Citation:
**Bosson, JB, Huss, M & Osipova E (2019). Dataset of the paper Disappearing World Heritage glaciers as a keystone of nature conservation in a changing system, Earth’s Future.**

